Understanding the Lomb-Scargle Periodogram

### 2016-12-27 PracticalLombScargle.tex
Add LS model figure

\citet{Scargle82} showed that while the classical periodogram of
\eq{classical-periodogram} has these desireable statistical properties
in the case of uniform sampling, this does not hold for
\eq{classical-periodogram} when the sampling becomes non-uniform.
To address this, he considered a generalized form of the periodogram,
\begin{equation}
  P(f) = \frac{A^2}{2}\left(\sum_n g_n \cos(2\pi f t_n)\right)^2
       + \frac{B^2}{2} \left(\sum_n g_n \sin(2\pi f t_n)\right)^2,
\end{equation}
where $A$ and $B$ are arbitrary functions of the frequency $f$ and
observing times $\{t_i\}$ (but not the values $\{g_n\}$, and showed
that you can choose one particular form of $A$ and $B$ such that
\begin{enumerate}
  \item The periodogram reduces to the classical form in the case of equally-spaced data,
  \item The periodogram has desireable statistical properties, in the sense discussed above,
  \item The periodogram is insensitive to global time-shifts in the data.
\end{enumerate}
The values of $A$ and $B$ leading to these properties lead to the following
modification of the classical expression for the periodogram:
\begin{eqnarray}
  P(f) =
  \frac{1}{2} \Bigg\{ &
  \bigg(\sum_n g_n \cos(2\pi f [t_n-\tau])\bigg)^2 \bigg/
  \sum_n \cos^2(2\pi f [t_n-\tau]) &\nonumber\\
  & + ~ \bigg(\sum_n g_n \sin(2\pi f [t_n-\tau])\bigg)^2 \bigg/
  \sum_n \sin^2(2\pi f [t_n-\tau]) & \Bigg\}
  \eqlabel{lomb-scargle-periodogram}
\end{eqnarray}
Here $\tau$ is defined for each $f$ to ensure time-shift invariance of $P(f)$:
\begin{equation}
  \tau = \frac{1}{4\pi f}\tan^{-1}\Bigg(
  \frac{\sum_n \sin(4\pi f t_n)}{\sum_n \cos(4\pi f t_n)}\Bigg).
  \eqlabel{tau-def}
\end{equation}
A remarkable feature of this result is that it is {\it identical} to the result
obtained by fitting a simple sinusoid to the data at each frequency $f$, and
constructing a ``periodogram'' from the $\chi^2$ goodness-of-fit at each
frequency--an estimator which was considered in some depth by \citet{Lomb76}.
In this view, the $\tau$ shift defined in \eq{tau-def} serves to orthogonalize
the normal equations used in the least squares analysis.

@@

Partly due to this deep connection between Fourier analysis and least-squares
analysis, the modified periodogram in \eq{lomb-scargle-periodogram}
has since become commonly referred to as the {\it Lomb-Scargle Periodogram}.
The Lomb-Scargle periodogram differs from the classical periodogram only to
the extent that the denominators $\sum_n \sin^2(2\pi f t_n)$ and
$\sum_n \cos^2(2\pi f t_n)$ differ from $N/2$, which is the expected value of
each of these quantities in the limit of very dense sampling.
Thus, in many cases of interest the Lomb-Scargle periodogram only differs
slightly from the classical/Schuster periodogram; an example of this is seen
in \fig{ls-comparison}.

Because of this similarity between the classical and Lomb-Scargle periodograms,
our discussion of windowing effects applies--at least qualitatively--to results
computed with the Lomb-Scargle periodogram.
In particular, reasoning about the effect of window functions on the observed
Lomb-Scargle power spectrum remains useful even if it is true in only a
qualitative sense.

\section{Using Lomb-Scargle Effectively}

\begin{itemize}
  \item Expected peak width
  \item How to choose the frequency grid
  \item Normalization and interpretation
\end{itemize}

@@

\section{Algorithmic Considerations}

\begin{itemize}
\item Press \& Rybicki method; show some benchmarks.
\item NFFT method
\end{itemize}

@@

\begin{itemize}
\item Uncertainties in Data
\item Bayesian view is (almost always) not useful
\item Multiterm makes the true period fit better, but also bumps the background noise.
\end{itemize}

@@

\appendix

\section{Non-uniform Nyquist}
A selection of quotations showing approaches to a non-uniform Nyquist limit:

\begin{description}

\item[Average Sampling Rate]

\citet{Scargle82} (citing the following: Beutler 1966, 1970; Masry \& Lui 1975; Higgins 1976; Wiley 1978; Gaster \& Roberts 1975, 1977; Kar, Hornkohl \& Farmer 1981; Ludeman 1981 \todo{read and summarize these!}):
\begin{quote}
Error-free recovery of a band-limited signal [i.e.~reproduction of the entire function $X(t)$ from the samples $X(t_i)$] can be achieved with irregular sampling as long as the mean sampling rate exceeds the Nyquist rate (i.e. the average number of samples per unit time must exceed twice the highest frequency component in the signal).
\end{quote}

\citet{NumRec}: 
\begin{quote}
One guide to choosing $f_{hi}$ is to compare it with the Nyquist frequency $f_c$ which would obtain if the $N$ data points were evenly spaced over the same span $T$, that is $f_c = N/(2T)$. The accompanying program includes an input parameter {\tt hifac}, defined as $f_{hi}/f_{c}$.
\end{quote}

\citet{Horne86}:
\begin{quote}
The largest frequency we calculated was $\pi N/T$ which
is the traditional Nyquist frequency for evenly-spaced data. The Nyquist
frequency is not well defined for unevenly spaced signals, but it can serve as
a reasonable upper limit for the calculation.
\end{quote}

\item[Harmonic Average of Sampling Rate]

\citet{Debosscher07}:
\begin{quote}
For the highest frequency, we used the average of the inverse time intervals
between the measurements: $f_N = 0.5(1/\Delta T)$ as a pseudo
Nyquist frequency. Note that $f_N$ is equal to the Nyquist frequency
in the case of equidistant sampling. For particular cases,
an even higher upper limit can be used \citep[see][]{Eyer99}.
Our upper limit should be seen as a compromise between
the required resolution to allow a good fitting, and computation
time.
\end{quote}

\item[Minimum Sample Spacing]

\citet{Percy86}:
\begin{quote}
... the Nyquist frequency is not well defined for input data obtained at unequally spaced time intervals, the usual case in astronomy \citep{Scargle82}.
Theoretically such a data set contains informatin on the periodicities down to
$\Delta t = \min(t_i - t{i-1})$. In practice, however, a pseudo-Nyquist frequency may be defined by averaging $\Delta t = (t_i - t_{i-1})$, where large, uncharacteristic temporal gaps are avoided. Alternatively, the harmonic mean of all $\Delta t$ may be used. The result is that a useful pseudo-Nyquist frequency may be defined by $\nu_{Ny} = (2 \langle\Delta t \rangle)^{-1}$.
\end{quote}

\citet{Press89}:
\begin{quote}
It is often meaningful to examine frequencies significantly higher than the
Nyquist frequency that would obtain if the same number of data points were
evenly spaced in the same total length of time. Some spectral information is
obtainable for frequencies all the way up to something like half the inverse
spacing of the {\it closest} spaced points.
\end{quote}

\citet{Roberts87}:
\begin{quote}
...for arbitrary $\{t_r\}$, the sampling theorem tells us nothing.
If the data samples are otherwise equally spaced but with missing
points, the theorem says that the data completely determine
a function whose FT is zero for $|v| > 1/(2\Delta_{max})$, where $\Delta_{max}$
is the {\it largest completely sampled} data spacing. However,
there are smaller spacings, and these certainly carry information
about frequencies greater than $1/(2\Delta_{max} )$; some information
is available about frequencies as high as
$1/(2\Delta_{min} )$, where $\Delta_{min}$ is the smallest spacing between data
points. Furthermore, if the $\{t_r\}$ are more or less randomly
distributed, so that a wide range ofspacings are present and
there is little redundancy in the spacing between various
points, tests have shown (Paper II) that significant information
is available on frequencies greater than $l/(2\Delta_{min})$.

Nonetheless, in the present paper we will restrict ourselves
to frequencies obeying
$\nu < \nu_{max} = 1/(2\Delta_{min})$.
\end{quote}

\citet{Hilditch01}:
\begin{quote}
The highest frequency, for equally spaced data, is formally given by the {\it Nyquist frequency}, $f_N = 1/(2\Delta t)$, where $\Delta t$ is the {\it sampling interval} in the data string.
For unequally spaced data, it is common practice to estimate a pseudo-Nyquist frequency from the minimum value of $\Delta t$.
\end{quote}

\item[Getting it Right]

\citet{ICVG2014}:
\begin{quote}
As a good choice for the maximum search frequency, a pseudo-Nyquist frequency $\omega_{max} = \pi/\Delta t$, where $1/\Delta t$ is the median of the inverse time interval between data points, was proposed by \citet{Debosscher07} (in the case of even sampling, $\omega_{max}$ is equal to the Nyquist frequency). In practice, this choice may be a gross underestimate because unevenly sampled data can detect periodicity with frequencies even higher than $2\pi / \Delta t_{min}$ \citep[see][]{Eyer99}. An appropriate choice of $\omega_{max}$ thus depends on sampling (the phase coverage at a given frequency is the relevant quantity) and needs to be carefully chosen: a hard limit on maximum detectable frequency is of course given by the time interval over which individual measurements are performed, such as imaging exposure time.
\end{quote}

\end{description}

>>>>

\citet{Scargle82} showed that while the classical periodogram of
\eq{classical-periodogram} has these desireable statistical properties
in the case of uniform sampling, this does not hold for
\eq{classical-periodogram} when the sampling becomes non-uniform.
To address this, he considered a generalized form of the periodogram,
\begin{equation}
  P(f) = \frac{A^2}{2}\left(\sum_n g_n \cos(2\pi f t_n)\right)^2
       + \frac{B^2}{2} \left(\sum_n g_n \sin(2\pi f t_n)\right)^2,
\end{equation}
where $A$ and $B$ are arbitrary functions of the frequency $f$ and
observing times $\{t_i\}$ (but not the values $\{g_n\}$, and showed
that you can choose one particular form of $A$ and $B$ such that
\begin{enumerate}
  \item The periodogram reduces to the classical form in the case of equally-spaced data,
  \item The periodogram has desireable statistical properties, in the sense discussed above,
  \item The periodogram is insensitive to global time-shifts in the data.
\end{enumerate}
The values of $A$ and $B$ leading to these properties lead to the following
modification of the classical expression for the periodogram:
\begin{eqnarray}
  P(f) =
  \frac{1}{2} \Bigg\{ &
  \bigg(\sum_n g_n \cos(2\pi f [t_n-\tau])\bigg)^2 \bigg/
  \sum_n \cos^2(2\pi f [t_n-\tau]) &\nonumber\\
  & + ~ \bigg(\sum_n g_n \sin(2\pi f [t_n-\tau])\bigg)^2 \bigg/
  \sum_n \sin^2(2\pi f [t_n-\tau]) & \Bigg\}
  \eqlabel{lomb-scargle-periodogram}
\end{eqnarray}
Here $\tau$ is defined for each $f$ to ensure time-shift invariance of $P(f)$:
\begin{equation}
  \tau = \frac{1}{4\pi f}\tan^{-1}\Bigg(
  \frac{\sum_n \sin(4\pi f t_n)}{\sum_n \cos(4\pi f t_n)}\Bigg).
  \eqlabel{tau-def}
\end{equation}
A remarkable feature of this result is that it is {\it identical} to the result
obtained by fitting a simple sinusoid to the data at each frequency $f$, and
constructing a ``periodogram'' from the $\chi^2$ goodness-of-fit at each
frequency--an estimator which was considered in some depth by \citet{Lomb76}.
In this view, the $\tau$ shift defined in \eq{tau-def} serves to orthogonalize
the normal equations used in the least squares analysis.
In other words, the Lomb-Scargle periodogram essentially {\it assumes a
sinusoidal model} for the data, which is visualized in \fig{ls-model} for
the data we had seen in \fig{LINEAR-data}.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig18_ls_model}
  \caption{The sinusoidal model implied by the Lomb-Scargle periodogram for
    the LINEAR data seen in \fig{LINEAR-data}.
    Although a sinusoid does not perfectly fit the data, it is close enough
    that it serves to find the correct period without a problem.
    \figlabel{ls-model}}
\end{figure}

Partly due to the deep connection between Fourier analysis and least-squares
analysis, the modified periodogram in \eq{lomb-scargle-periodogram}
has since become commonly referred to as the {\it Lomb-Scargle Periodogram}.
The Lomb-Scargle periodogram differs from the classical periodogram only to
the extent that the denominators $\sum_n \sin^2(2\pi f t_n)$ and
$\sum_n \cos^2(2\pi f t_n)$ differ from $N/2$, which is the expected value of
each of these quantities in the limit of very dense sampling.
Thus, in many cases of interest the Lomb-Scargle periodogram only differs
slightly from the classical/Schuster periodogram; an example of this is seen
in \fig{ls-comparison}.

@@

Because of this similarity between the classical and Lomb-Scargle periodograms,
our discussion of windowing effects applies---at least qualitatively---to
results computed with the Lomb-Scargle periodogram.
In particular, reasoning about the effect of window functions on the observed
Lomb-Scargle power spectrum remains useful even if it is true in only a
qualitative sense.

\section{Practical Considerations when using Lomb-Scargle Periodograms}

\subsection{Choosing a Frequency Grid}

\subsection{Normalization of the Periodogram}

\subsection{Algorithmic Considerations}

@@

@@

\begin{itemize}
\item floating-mean
\item Uncertainties in Data
\item Bayesian view is (almost always) not useful
\item Multiterm makes the true period fit better, but also bumps the background noise.
\end{itemize}

@@
